---
title: "confederate monuments analysis"
author: "Sahana Jayaraman, Tori Bergel, Theresa Cotton, and Manuela Lopez Restrepo" 
date: "12/13/2020"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The purpose of this document is to display the analysis behind the five findings that generated the story group 2 pitched for our final. Each finding is a header, with the code that engineers that finding listed below with line-by-line descriptions. Using the Southern Poverty Law Center's Whose Heritage dataset, our story centers on 3 trees in California, all named after Robert E. Lee. 

##Load Libraries 
```{r}
# install.packages("leaflet")
library(tidyverse)
library(janitor)
library(leaflet)

```

##Loading in the data 
```{r}
#reading in the data and converting the coordinates column into two separate numeric columns for latitude and longitude so leaflet can map
heritage<- read_csv("data/whose_heritage.csv")%>%
  clean_names()%>%
  separate(coordinates, into = c("lat", "lng"),sep =",")%>%
  mutate(lat = as.numeric(lat), 
         lng = as.numeric(lng))

# Reconfiguring data frame so that each state falls into one category "South" or "North" or "Not Applicable" to better view the number of monuments that fall into each category (monuments in union states v. monuments in confederate states)
side<-heritage%>%
  rename(clean_side = side)%>%
  group_by(state, clean_side)%>%
  count()%>%
  filter(!is.na(clean_side))%>%
  arrange(state)%>%
  mutate(clean_side = case_when(
    state =="AL" ~ "South", 
    state =="AR" ~ "South",
    state == "TX" ~ "South",
    state == "VA" ~ "South",
    TRUE ~ clean_side
  ))%>%
  select(-n)%>%
  distinct()


# Join cleaned side with heritage 
heritage<- heritage%>%
  inner_join(side, by = c("state"))%>%
  select(-side)%>%
  rename(side = clean_side)


```
## Finding #1: Northern states, while they have less monuments than southern states, have a higher percentage of symbols removed. 
```{r}
#Q: How many monuments by side? 
#Start with Heritage dataset AND THEN 
#group by side AND THEN
#Count the number of monuments in each side AND THEN
#Arrange the list so the side with the most monuments is at the top.
heritage%>%
  group_by(side)%>%
  count()%>%
  arrange(desc(n))

#total number of monuments: Union (northern) states = 46, Confederate (southern) states = 1757, border states = 98, "not applicable" states = 39

#Q: What percentage of monuments removed per side? 
#Create a new object called percent removed, feeding it the heritage dataset AND THEN
#Group by side, and tracking status AND THEN 
#Recode variables so that symbols where statues were removed but pedestals remain and renamed symbols show up as "removed" AND THEN
#count the number of symbols in each category AND THEN
#pivot the table wider to divide removed symbols by live symbols AND THEN
#clean names to make manipulation easier AND THEN 
#create a new column called percent_removed, and divide the new 'removed' column values by the 'live' column values to geth the percentage of symbols removed in each side


percent_removed<-heritage%>%
  group_by(side, tracking_status)%>%
  mutate(tracking_status = case_when(
    tracking_status =="Live (Statue Removed; Pedestal Remains)" ~ "Removed", 
    tracking_status =="Renamed" ~ "Removed", 
    TRUE ~ tracking_status
  ))%>%
  count()%>%
  pivot_wider(names_from = tracking_status, values_from = n)%>%
  clean_names()%>%
  mutate(percent_removed = (removed / live)*100)
  


percent_removed

# Shows the percent removed in each category - for the purpose of this exercise, we counted "renamed" and "statue removed" under the umbrella term "removed." Percentage of monuments removed in northern states is much higher than in southern states; southern states have relocated a notable number of monuments and the status on 4 are pending. 
```


##Finding #2: California is the Union state with the most confederate symbols
```{r}

# Q: Which Northern(Union) state has the most confederate monuments?
#creating a new object called "northern_states" and feeding it the heritage dataset AND THEN
#filtering to only return rows that contain value "North" in the side column AND THEN
#grouping by state AND THEN 
#counting the number of symbols in each northern state AND THEN
#arranging the list so that the state with the most confederate monuments is at the top

northern_states<- heritage%>%
  filter (side =="North")%>%
  group_by(state)%>%
  summarise(count = n())%>%
  arrange(desc(count))

northern_states

# A: California has the most monuments — so we're focusing in on Cali 
```



##Finding #3:Though nearly half of the monuments in California have been removed since 2016, the data says three trees remain 
```{r}
# Q: How many monuments in Cali have come down?
#create new object called 'cali_monuments_down' and feed it the heritage dataset AND THEN
#filter state to return rows that have value "CA" in the state column AND THEN
#filter tracking status to return objects removed and renamed AND THEN
#group objects by year removed AND THEN
#count the number of removed/renamed California monuments in each year group  
cali_monuments_down <- heritage %>%
  filter(state == "CA") %>%
  filter(tracking_status %in% c("Removed", "Renamed"))%>%
  group_by(year_removed)%>%
  count()
 
cali_monuments_down

# A: 6, starting in 2016 and the latest removed this year. Cali has 14 total, so nearly half have been removed 

# Q: What is still live? 
# create object 'cali_what' feed it heritage AND THEN 
# filter to return live monuments in California AND THEN 
# group by symbols by type and category AND THEN
# count symbols in categories AND THEN 
# Arrange so that category with most symbols is at top of list

cali_what <-heritage %>%
  filter(state== "CA", tracking_status == "Live")%>%
  group_by(symbol_type, symbol_category)%>%
  count()%>%
  arrange(desc(n))
cali_what

# A: The most still-live monuments are parks/trails, 

# Q: Exploring parks/trails:
# take the heritage data-frame AND THEN
# Filter to return Live California monuments that are parks/trails 

heritage%>%
  filter(state== "CA", tracking_status == "Live", symbol_category == "Parks/Trails")

# A: all the parks/trails symbols remaining are trees
```



##Finding #4: All three of these trees are named after Robert E. Lee, that they are all sequoias, one named in 1875, one named much later in 1937 and one unknown, and theyre all located in national parks. 
```{r}

# take the heritage dataframe AND THEN
# Filter to return Live California monuments that are parks/trails 
heritage%>%
  filter(state== "CA", tracking_status == "Live", symbol_category == "Parks/Trails")
#All three live monuments in the parks/trails category  are trees
#Dedicated or named in 1875 and 1937 (missing data for Yosemite National Park tree)
#All same type of tree (Sequoia)
#Trees have not been removed or renamed since their dedication (live tracking status)
#Sponsors are different (One is Richard Field and one is United Daughters of the Confederacy and one is NA)
#The three trees are located in Sequoia National Park, Kings Canyon National Park, and Yosemite National Park and they're all dedicated to Robert E. Lee.

# Q: viewing trees on a map
# create new object three_trees and feed in dataset 'heritage' AND THEN
# filter to return the three trees AND THEN 
# Recode variables in column "feature_name" so all the trees are labeled "tree"
three_trees<-heritage%>%
  filter(symbol_type == "Other", state == "CA", honorees == "Robert E. Lee" )%>%
  mutate( feature_name = case_when(
    feature_name == "Robert E. Lee Sequoia Tree at Sequoia National Park" ~ "tree", 
    feature_name == "Robert E. Lee Sequoia Tree at Kings Canyon National Park" ~ "tree",
      feature_name == "Robert E. Lee Sequoia Tree at Yosemite National Park" ~ "tree",
    ))

#view three_trees
three_trees

three_trees%>%
  leaflet()%>%
  addProviderTiles(provider = "CartoDB.Positron")%>%
  addMarkers(lat = ~lat, lng = ~lng, 
             label = ~unique_id)

  
#Note: Two of the trees are missing lat/lng values and can't be mapped, so they don't show up.
```


## FInding #5: That there are trees in other states named after confederate generals, but it appears that california has the most. 
```{r}

# Create  object called 'american_trees' and feed in heritage dataset AND THEN 
# recoding the feature names of oaks and other trees to show up as "tree" AND THEN 
# filtering to return only recoded variables AND THEN
# grouping by state AND THEN
# counting the number of trees per state
american_trees<-heritage%>%
  mutate( feature_name = case_when(
    feature_name == "Robert E. Lee Memorial Tree" ~ "tree",
     feature_name == "Robert E. Lee Sequoia Tree at Sequoia National Park" ~ "tree", 
    feature_name == "Robert E. Lee Sequoia Tree at Kings Canyon National Park" ~ "tree",
    feature_name == "Robert E. Lee Sequoia Tree at Yosemite National Park" ~ "tree",
    feature_name == "Confederate Oak" ~ "tree",
    feature_name == "Lee Oak (Tree)" ~ "tree",
    TRUE ~ feature_name
    ))%>%
  filter( feature_name == "tree")%>%
  group_by(state)%>%
  count()

american_trees

# A: There are other trees in other states named after confederate soldiers/generals, but California appears to have the most 
```
